---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>Statistics and Data Visualizations in Climate Science</h1></center>
<center><h1>with R and Python</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book By</h3></center>
<center><h3>SSP Shen and GR North</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">Version 1.0 released in July 2023 and coded by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>

## Chapter 2: Elementary Probability and Statistics

### R Plot Fig. 2.1
```{r}
# Go to your working directory
# setwd("/Users/sshen/climstats") 
nycDat=read.csv("data/NYCDailyDatasummer2019.csv", header=T)
dim(nycDat)
#[1] 7555   13
dayNum=92
nycP=nycDat[1:dayNum, c(3,6)]

dw=c()
for (i in 1:dayNum){if (nycP[i,2] >= 0.2){dw[i] = 1} 
  else {dw[i]=0} }
dw
n0=which(dw==0)
n1=which(dw==1)
m=dayNum - 1
par(mfrow=c(1,1))
par(mar=c(2.5,4.5,2.5,4))
plot(1:dayNum, nycP[,2], type="s",
     main="Daily Precipitation of New York City: 
      1 Jun 2019- 31 Aug 2019",
     ylab="Precipitation [mm/day]", 
     xaxt="n", xlab="",
     cex.lab=1.4, cex.axis=1.4)
axis(1, at=c(1, 30, 61, 92), 
     labels=c("1 June", "1 July", "1 Aug", "31 Aug"),
     cex.axis=1.4)
par(new=TRUE)
plot(n0+1.0, dw[n0]-0.2, cex.lab=1.4,
     ylim=c(0,15), pch=15, cex=0.4, col="red",
     axes=FALSE, xlab="", ylab="")
points(n1+1, dw[n1], pch=15, col="blue", cex=0.4)
axis(4, col="blue", col.axis="blue", 
     at=c(0,1), labels =c("D", "W"), 
     las=2, cex.axis=1)
mtext("Dry or Wet Days", 
      col="blue", side=4, line=1.5, cex=1.4)

```


<p>&nbsp;</p>
### R Plot Fig. 2.2
```{r}
p=0.6471
n=1:12
pdfn= (1-p)*p^n
par(mar=c(4.5,4.5,3,0.5))
plot(n, pdfn, 
     ylim=c(0,0.25), type="h", lwd=15,
     cex.lab=1.5, cex.axis=1.5,
     xlab="Length of Dry Spell: n", ylab="Probability",
     main="Probability Distribution Function of Dry Spell")
text(1.5,0.25, "(a)",cex=1.5)

p=0.6471
N=1:12
cdfN= 1-p^N
par(mar=c(4.5,4.5,3,0.5))
plot(N, cdfN, 
     ylim=c(0,1), type="s", lwd=3,
     cex.lab=1.5, cex.axis=1.5,
     xlab="N: Dry Spell Not Longer Than N Days", 
     ylab="Cumulative Probability",
     main="Cumulative Distribution Function of Dry Spell")
text(1.5,0.95, "(b)",cex=1.5)
```


<p>&nbsp;</p>
### R Plot Fig. 2.3
```{r}
n=20
x=0:n
pdfdx=dbinom(x, size=n, prob=0.3)
par(mar=c(4.5,4.5,2,4.5))
plot(x, pdfdx, type="h", lty=1, lwd=3,
     xlab="Number of successes: k", ylab="Probability", 
     main="Binomial Distribution: n=20, p=0.3",
     cex.lab=1.4, cex.axis=1.4)
text(11, 0.05, "PDF", cex=2)
par(new=TRUE)
cdfx=pbinom(x, size=n, prob=0.3)
plot(x, cdfx, col="blue",
     type="o", lty=2, axes=FALSE,
     xlab="", ylab="", cex.axis=1.4)
axis(4, col="blue", col.axis = "blue", cex.axis=1.4)
mtext("Cumulative Probability", col="blue",
      side=4, line=3, cex=1.4)
text(11, 0.9, "CDF", cex=2, col="blue")
```


<p>&nbsp;</p>
### R Plot Fig. 2.3: Save as an eps file
```{r}
#setwd("/Users/sshen/climstats")
#setEPS() # save the .eps figure file to the working directory
#postscript("fig0203.eps", width = 7, height = 5)
n=20
x=0:n
pdfdx=dbinom(x, size=n, prob=0.3)
par(mar=c(4.5,4.5,2,4.5))
plot(x, pdfdx, type="h", lty=1, lwd=3,
     xlab="Number of successes: k", ylab="Probability", 
     main="Binomial Distribution: n=20, p=0.3",
     cex.lab=1.4, cex.axis=1.4)
text(11, 0.05, "PDF", cex=2)
par(new=TRUE)
cdfx=pbinom(x, size=n, prob=0.3)
plot(x, cdfx, col="blue",
     type="o", lty=2, axes=FALSE,
     xlab="", ylab="", cex.axis=1.4)
axis(4, col="blue", col.axis = "blue", cex.axis=1.4)
mtext("Cumulative Probability", col="blue",
      side=4, line=3, cex=1.4)
text(11, 0.9, "CDF", cex=2, col="blue")
#dev.off()
```


<p>&nbsp;</p>
### Plot Fig. 2.4
```{r}
x=seq(-5,5, len=101)
y=dnorm(x)
plot(x, y, 
     type="l", lty=1, 
     xlab="x", ylab="Probability Density f(x)", 
     main="Standard Normal Distribution",
     cex.lab=1.4, cex.axis=1.4)
lines(x,rep(0, 101))
xx=c(x[60], x[60], x[70], x[70])
yy=c(0, y[60], y[70], 0)
polygon(xx, yy, col="brown") 
text(0.7, 0.01, "a", col="brown")
text(2.1, 0.01, "b", col="brown")
text(1.7,0.2, expression(f(x)), cex=1.4)
par(new=TRUE)
plot(x, pnorm(x), type="l", lty=2, col="blue",
     axes=FALSE, ylab="", xlab="", cex.axis=1.4)
axis(4, col="blue", col.axis = "blue", cex.axis=1.4)
text(3,0.9, expression(F(x)), cex=1.4, col="blue")
mtext("Cumulative Probability F(x)", col="blue",
      side=4, line=3, cex=1.4)
text(3.3,0.3, cex=1.3, col="brown",
     expression(P[ab]==integral(f(x)*dx, a,b)))
```


<p>&nbsp;</p>
### R plot Fig. 2.5: 2023 version
```{r}
par(mar=c(4.5, 4.5, 2.5, 4.5))
x<- rnorm(200, mean=5, sd=3)
h<-hist(x, breaks=15, col="gray", 
        xlim=c(-5,15), ylim=c(0,30),
        xlab=expression("Random numbers x of N(5," * 3^2 *")"),
        ylab="Frequency",
        main="Histogram and Its Normal PDF Fit", 
        cex.lab=1.4, cex.axis=1.4)
xfit<-seq(min(x),max(x),length=40)
yfit<-dnorm(xfit,mean=mean(x),sd=sd(x))
#yfit <- yfit*diff(h$mids[1:2])*length(x)
yfit <- yfit
#diff(h$mids[1:2])*length(x) =200 
#is the total histogram area
points(x, rep(0, 200), cex=0.8, pch=3, col="blue")
par(new = TRUE)
plot(xfit, yfit, type = 'l', lwd = 3, axes = FALSE,
     xlab = "", ylab = "", col = 'red')
axis(4, col='red', col.axis="red", cex.axis =1.4)
mtext("Density", side =4, line = 3, cex = 1.4,
      col = 'red')
```


<p>&nbsp;</p>
### Plot Fig. 2.5: The book version
```{r}
par(mar=c(4.5, 4.5, 2.5, 0.5))
x<- rnorm(200, mean=5, sd=3)
h<-hist(x, breaks=15, col="gray", 
        xlim=c(-5,15), ylim=c(0,30),
        xlab=expression("Random numbers x of N(5," * 3^2 *")"),
        ylab="Frequency/Density",
        main="Histogram and Its Normal Curve Fit", 
        cex.lab=1.4, cex.axis=1.4)
xfit<-seq(min(x),max(x),length=40)
yfit<-dnorm(xfit,mean=mean(x),sd=sd(x))
yfit <- yfit*diff(h$mids[1:2])*length(x)
#diff(h$mids[1:2])*length(x) =200 
#is the total histogram area
points(x, rep(0, 200), cex=0.8, pch=3, col="blue")
lines(xfit, yfit,  lwd=2)
```


<p>&nbsp;</p>
### Plot Fig. 2.6
```{r}
#install.packages("ggExtra") 
library(ggplot2)
df <- data.frame(x = rnorm(1000, 5, 3), y = runif(1000, 0, 5))
p <- ggplot(df, aes(x, y)) + geom_point() + theme_classic()
ggExtra::ggMarginal(p, type = "histogram")

```


<p>&nbsp;</p>
### Plot Fig. 2.7: R code for Poisson Distribution
```{r}
k = 0:20
y = dpois(k, lambda = 5)
par(mar=c(4.5,4.5,2,4.5))
plot(k, y, 
     type="p", lty=1, xlab="k", ylab="", 
     main="Poisson Distribution: Mean Rate = 5",
     cex.lab=1.4, cex.axis=1.4)
mtext(side=2, line=3, cex=1.4, 'Probability f(k)')
par(new=TRUE)
plot(k, ppois(k, lambda = 5), type="p", pch=16, 
     lty=2, col="blue",
     axes=FALSE, ylab="", xlab="", cex.axis=1.4)
axis(4, col="blue", col.axis = "blue", cex.axis=1.4)
mtext("Cumulative Probability F(k)", col="blue",
      side=4, line=3, cex=1.4)

##
dpois(10, lambda=5)
```

<p>&nbsp;</p>
### Plot Fig. 2.8: R code
```{r}
x=seq(0,20, len=201)
ycauchy=dcauchy(x, location = 10, scale=1)
ygauss=dnorm(x, mean=10, sd=sqrt(pi/2))
plot(x, ycauchy, type="l", lwd=2, 
     ylab="Density", 
     main="Comparison between Cauchy and Gaussian Distributions",
     cex.lab=1.4, cex.axis=1.4)
legend(-1,0.35, legend="Cauchy distribution", 
       cex=1.2, bty="n", lty=1, lwd=2)
lines(x, ygauss, lty=2)
legend(-1,0.32, legend="Normal distribution", 
       cex=1.2, bty="n", lty=2)


##
x= seq(1,10, by=0.2)
plot(x,dchisq(x, df = 4))
```


<p>&nbsp;</p>
### Plot Fig. 2.9: R code
```{r}
#setwd("/Users/sshen/climstats")
da1 =read.csv("data/EdmontonT.csv", header=TRUE)
x=da1[,3]
m1 = mean(x)
s1 = sd(x)
xa = (x- m1)/s1
y = matrix(xa[1:1632], ncol=6, byrow=TRUE)
w = rowSums(y^2)
hist(w, breaks= seq(0,40,by=1),
     xlim=c(0,40), ylim=c(0,50),
     main="Histogram and its Chi-square Fit
for Edmonton temperature data", 
     xlab=expression("Chi-square data ["~degree~C~"]"^2),
     cex.lab=1.4, cex.axis=1.4)
par(new=TRUE)
density = function(x) dchisq(x, df=6)
x=seq(0,40, by=0.1)
plot(x, density(x), type="l", lwd=3,
     col="blue", ylim=c(0,0.15),
     axes=FALSE, bty="n", xlab="", ylab="")
axis(4, cex.axis=1.4, col='blue', col.axis='blue')
mtext("Chi-square Density", cex=1.4,
      side = 4, line = 3, col="blue")
text(20, 0.1, cex=1.4, col="blue",
     "Chi-square distribution: df=6")
```


<p>&nbsp;</p>
### Lognormal Distribution of the June Madison Precipitation
```{r}
#setwd("/Users/sshen/climstats")
Madison = read.csv("data/MadisonP.csv", header=TRUE)
Madison[1:2,]#Take a look at the first two lines of the data
daP = matrix(Madison[,7], ncol=12, byrow=TRUE)
x = daP[,6] #The June precipitation data
n=length(x)
m1=mean(x)
A = log(m1)- sum(log(x))/n
c = (1/(4*A))*(1 + sqrt(1 + 4*A/3))
beta = c/m1
mu=mean(log(x))
sig = sd(log(x))
round(c(mu, sig), digits = 2)
#[1] 4.52 0.65
modelmean = exp(mu + 0.5*sig^2)
modelsd = sqrt((exp(sig^2)-1)*exp(2*mu + sig^2))
datamean = mean(x)
datasd = sd(x)
round(c(modelmean, datamean, modelsd, datasd), digits =2)
#[1] 112.81 109.79  81.26  63.94
```

