---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>Statistics and Data Visualizations in Climate Science</h1></center>
<center><h1>with R and Python</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press Book by</h3></center>
<center><h3>SSP Shen and GR North</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">Version 1.0 released in July 2023 and coded by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>

## Chapter 6: Covariance Matrices, EOFs, and PCs

### Plot Fig. 6.1a, b, c: and Covariance Map by R code
```{r}
#Covariance from NOAAGlaobalTemp data: December SAT anomalies  
#setwd('/Users/sshen/climstats') 
#install.packages('ncdf4')
library(ncdf4)
nc=ncdf4::nc_open("data/air.mon.anom.nc") #Read data
nc #Check metadata of the dataset
nc$dim$lon$vals # 2.5 - 357.5
nc$dim$lat$vals # -87.5 - 87.5
nc$dim$time$vals
lon <- ncvar_get(nc, "lon")
lat <- ncvar_get(nc, "lat")
time<- ncvar_get(nc, "time")
library(chron)
month.day.year(29219,c(month = 1, day = 1, year = 1800))
#1880-01-01
sat<- ncvar_get(nc, "air")
dim(sat)
#[1] 72   36 1674 
#1674 months=1880-01 to 2019-06, 139 years 6 mons

Dec = seq(12, 1674, by=12)
Decsat=sat[,, Dec]
N = 72*36
P = length(Dec)
STsat = matrix(0, nrow=N, ncol=P)
for (k in 1:P){STsat[,k]=as.vector(Decsat[,,k])}
colnames(STsat)<-1880:2018
STsat[1:4,1:4]
LAT=rep(lat, each=72)
LON=rep(lon,36)
STanom=cbind(LAT, LON, STsat)
dim(STanom)
#[1] 2592  141

#Plot Fig. 6.1a: Zonal covariance matrix 
#Select only the data for the equatorial band -2.5S
n1<-which(STanom[,1]>-4&STanom[,1]<0
          &STanom[,2]>0&STanom[,2]<360)
dim(STanom)
#[1] 2592  141
length(n1)
#[1] 72 longitude grid boxes at -2.5S
P1=84
P2=133
P= P2 - P1 + 1
dat1=STanom[n1,P1:P2] #1961-2010
dim(dat1)
#[1] 72 50
#72 grid boxes, 50 years of Dec from 1961-2010. 
dat1[1:3,48:50]
Lat1=STanom[n1,1]
Lon1=STanom[n1,2]
AreaFac = sqrt(cos(Lat1*pi/180))
dat2 = dat1 - rowMeans(dat1) #Minus the mean
Banddat = AreaFac*dat2
covBand = Banddat%*%t(Banddat)
max(covBand)
#[1] 90.67199
min(covBand)
#[1] -6.191734
int=seq(-7, 92,length.out=81)
rgb.palette=colorRampPalette(c('blue', 'darkgreen',
 'green', 'yellow','pink','red','maroon'),
                             interpolate='spline')
#setEPS() #Plot the figure and save the file
#postscript("fig0601a.eps", width = 7, height = 5.5)
par(mar=c(4.2,5.0,1.8,0.0))
par(cex.axis=1.8,cex.lab=1.8, cex.main=1.7)
ticks = c(60,  180, 300)
filled.contour(Lon1, Lon1, covBand, 
               color.palette=rgb.palette, levels=int,
               plot.title=title(main=
 expression("Zonal SAT Covariance: 2.5"*degree*S),
    xlab="Longitude", ylab="Longitude"),
   plot.axes = { axis(1, at= ticks); axis(2, at= ticks, las = 0)},
            key.title=title(main=expression("["*degree*"C]"^2))
      )
#dev.off()

###variance as the diagonal covariance
varzonal = diag(covBand)/(AreaFac)^2
plot(Lon1, varzonal, type = 'h', lwd =2)

#Plot Fig. 6.1b: Meridional covariance matrix
#Select only the data for the meridional band 2.5E
n1<-which(STanom[,1]>-90&STanom[,1]<90
          &STanom[,2]>0&STanom[,2]<4)
P1=84 #time index for Dec 1961
P2=133 #Dec 2010
P= P2 - P1 + 1 #=50 Decembers
dat1=STanom[n1,P1:P2] #1961-2010
Lat1=STanom[n1,1]
Lon1=STanom[n1,2]
AreaFac = sqrt(cos(Lat1*pi/180)) 
dat2 = dat1 - rowMeans(dat1)
Banddat = AreaFac*dat2
covBand = Banddat%*%t(Banddat)
max(covBand, na.rm=TRUE)
#[1] 115.6266
min(covBand, na.rm=TRUE)
#[1] -33.72083
int=seq(-35,120,length.out=81)
rgb.palette=colorRampPalette(c('blue', 'darkgreen','green', 
                               'yellow','pink','red','maroon'),
                             interpolate='spline')
ticks = seq(-90, 90, by=30)
#setEPS() #Plot the figure and save the file
#postscript("fig0601b.eps", width = 7, height = 5.5)
par(mar=c(4.2,5.0,1.8,0.0))
par(cex.axis=1.8, cex.lab=1.8, cex.main=1.7)
filled.contour(Lat1, Lat1, covBand, 
               color.palette=rgb.palette, levels=int,
               plot.title=title(main=
      expression("SAT Meridional Covariance: 2.5"*degree*E),
                    xlab="Latitude", ylab="Latitude"),
plot.axes={axis(1, at=ticks); axis(2, at=ticks, las =0); 
           grid()},
      key.title=title(main=expression("["*degree*"C]"^2)))     
#dev.off()

#plot Fig. 6.1(c)
month.day.year(time[1416],c(month = 1, day = 1, year = 1800))
#Dec 1997
mapmat= sat[,,1416]
mapmat=pmax(pmin(mapmat,5),-5)
int=seq(-5, 5,length.out=51)
rgb.palette=colorRampPalette(c('black','blue',
  'darkgreen','green', 'white','yellow','pink',
  'red','maroon'), interpolate='spline')
#setEPS() #Plot the figure and save the file
#postscript("fig0601c.eps", width = 7, height = 3.5)
par(mar=c(4.2,5.0,1.8,0.0))
par(cex.axis=0.9,cex.lab=0.9, cex.main=0.8)
library(maps)
filled.contour(lon, lat, mapmat, 
               color.palette=rgb.palette, levels=int,
          plot.title=title(main="December 1997 SAT Anomalies",
                          xlab="Longitude",ylab="Latitude"),
               plot.axes={axis(1); axis(2); 
                 map('world2', add=TRUE);grid()},
               key.title=title(main=expression(degree*"C")))
segments(x0=-20,y0=-2.5, x1=255, y1=-2.5, lwd = 5, lty = 2)
segments(x0=-15,y0=-90, x1=-15, y1=90, lwd = 5, lty = 2)
#dev.off()

#
#
#Zonal covariance matrix 
#Select only the data for the equatorial band -2.5S
n1<-which(STanom[,1]>-4&STanom[,1]<0
          &STanom[,2]>0&STanom[,2]<360)
dim(STanom)
#[1] 2592  142
length(n1)
#[1] 72 longitude grid boxes at -2.5S
P1=84
P2=133
P= P2 - P1 + 1
dat1=STanom[n1,P1:P2] #1961-2010
dim(dat1)
#[1] 72 50
#72 grid boxes, 50 years of Dec from 1961-2010. 
dat1[1:3,47:50]
Lat1=STanom[n1,1]
Lon1=STanom[n1,2]
AreaFac = sqrt(cos(Lat1*pi/180))
dat2 = dat1 - rowMeans(dat1) #Minus the mean
Banddat = AreaFac*dat2
covBand = Banddat%*%t(Banddat)
```


<p>&nbsp;</p>
### R plot 6.2: Scree plot
```{r}
K = 10
eigCov =eigen(covBand) 
#covBand is for equatorial zonal band in Fig 6.1(a)
lam = eigCov$values
lamK=lam[1:K]
#setEPS() #Plot the figure and save the file
#postscript("fig0602.eps", width = 6, height = 4)
par(mar=c(4,4,2,4), mgp=c(2.2,0.7,0))
plot(1:K, 100*lamK/sum(lam), ylim=c(0,80), type="o", 
     ylab="Percentage of Variance [%]",
     xlab="EOF Mode Number", 
     cex.lab=1.2, cex.axis = 1.1, lwd=2, 
     main="Scree Plot of the First 10 Eigenvalues")
legend(3,30, col=c("black"),lty=1, lwd=2.0,
       legend=c("Percentange Variance"),bty="n",
       text.font=2,cex=1.0, text.col="black")
par(new=TRUE)
plot(1:K,cumsum(100*lamK/sum(lam)),
     ylim = c(60,100), type="o",
     col="blue",lwd=2, axes=FALSE,
     xlab="",ylab="")
legend(3,80, col=c("blue"),lty=1,lwd=2.0,
       legend=c("Cumulative Percentage Variance"),bty="n",
       text.font=2,cex=1.0, text.col="blue")
axis(4, col="blue", col.axis="blue", mgp=c(3,0.7,0))
mtext("Cumulative Variance [%]",col="blue", 
      cex=1.2, side=4,line=2)
#dev.off()
```



<p>&nbsp;</p>
### Generate a Random Space-Time Field 
```{r}
#Step 1: Generate EOFs
N <- 100 #Number of spatial points
eof  <-  function(n, x) (sin(n*x)/sqrt(pi/2))*sqrt((pi/N))
x  <-  seq(0,pi, len=N)
sum(eof(4,x)^2)
#[1] 0.99  #Verify the normalization condition
sum(eof(1,x)*eof(2,x))
#[1] 3.035766e-18  #Verify orthogonality 

#Step 2: Generate PCs 
#install.packages('mvtnorm')
library(mvtnorm) #Multivariate normal
Mode <- 5
Ms <- 1000
univar <- diag(rep(1,Mode))
zeromean <- rep(0, Mode)
rs <- rmvnorm(Ms, zeromean, univar)
pcm=matrix(0, nrow=Ms, ncol=Mode)
for(m in 1:Mode){pcm[,m] = rs[,m]*sqrt(1/Ms)}
t(pcm[,1])%*%pcm[,1]
#1.010333 #Approximately normalized
t(pcm[,1])%*%pcm[,2]
#0.008040772 #Approximately independent/orthorgonal

#Step 3: Generate an independent random field
lam=c(10, 9, 4.1, 4, 2)
sqrlam = diag(sqrt(lam))
eofm = matrix(0, nrow=N, ncol=Mode)
for(m in 1:Mode){eofm[,m]=eof(m,x)}
Yxr  <-  eofm%*%sqrlam%*%t(pcm)
dim(Yxr)
#[1] 100 1000
```


<p>&nbsp;</p>
### Durbin-Watson (DW) Test for No Serial Correlation
```{r warning=FALSE}
#install.packages('lmtest')
library(lmtest)
Ms = 1000
r = 1:Ms
regYxr = lm(Yxr[10,] ~ r)
dwtest(regYxr)
#DW = 2.0344, p-value = 0.696
#Implying no significant serial correlation
```


<p>&nbsp;</p>
### Plot Fig. 6.3: R code
```{r}
#The first 100 samples of the generated field
r=1:100
#setEPS()
#postscript("fig0603.eps", width = 6, height = 4)
par(mar=c(3.5,3.5,2,0), mgp=c(2.0,0.7,0))
rgb.palette=colorRampPalette(
  c('black','blue','green', 
    'yellow','pink','red','maroon'),
  interpolate='spline')
filled.contour(r,x, t(Yxr[,1:100]), 
         color=rgb.palette, xlab="r", ylab="x", cex.lab=1.2,
          main="A random field realization from given EOFs",
          plot.axes={axis(1, cex.axis =1.1); 
            axis(2, las = 0, cex.axis= 1.2); grid()},
          key.title={par(cex.main=0.9); title(main="Value")}
)
#dev.off()
```


<p>&nbsp;</p>
### Scree Plot
```{r}
Mode = 1:5
lam=c(10, 9, 4.1, 4, 2)
samp = rep("S100", 5)
sd = sqrt(2/100)*lam
sd2 = sqrt(2/1000)*lam
par(mar=c(4.5, 4.5, 1, 0.5))
plot(Mode, lam, type="l", ylim = c(0,12),
     xlab="Mode", 
     ylab=expression("Variance   "*lambda),
     cex.lab=1.4, cex.axis=1.4)
points(Mode,lam + sd, pch="-", col="red", cex=2)
points(Mode,lam - sd, pch="-",col="red", cex=2)
segments(Mode,lam + sd, Mode,lam - sd, col="red")
points(Mode+0.06,lam + sd2, pch="-", col="blue", cex=2)
points(Mode+0.06,lam - sd2, pch="-",col="blue", cex=2)
segments(Mode +0.06,lam + sd2, 
         Mode + 0.06,lam - sd2, col="blue")
text(3.5,12, 
     "Red standard error bar: 100 samples", col="red")
text(3.5,11, 
     "Blue standard error bar: 1000 samples", col="blue")
```


<p>&nbsp;</p>
### Plot Fig. 6.4: Scree Plot by R code
```{r}
Mode = 1:5
lam=c(10.0, 9.0, 4.1, 4.0, 2.0)
samp = rep("S100", 5)
sd = sqrt(2/100)*lam
sd2 = sqrt(2/1000)*lam
#par(mar=c(4.5, 4.5, 1, 0.5))
#setEPS()
#postscript("fig0604.eps", width = 10, height = 8)
par(mar=c(3.5, 4.5, 0.2, 0.5), mgp=c(2.5, 1.0,0))
plot(Mode, lam, type="l", ylim = c(0,12),
     xlab="Mode", 
     ylab=expression("Variance   "*lambda),
     cex.lab=1.8, cex.axis=1.8)
points(Mode,lam + sd, pch="-", col="red", cex=2)
points(Mode,lam - sd, pch="-",col="red", cex=2)
segments(Mode,lam + sd, Mode,lam - sd, col="red")
points(Mode+0.06,lam + sd2, pch="-", col="blue", cex=2)
points(Mode+0.06,lam - sd2, pch="-",col="blue", cex=2)
segments(Mode +0.06,lam + sd2, Mode + 0.06,lam - sd2, 
         col="blue")
text(3.45,12, "Red standard error bar: 100 samples", 
         cex = 1.5, col="red")
text(3.5,11, "Blue standard error bar: 1000 samples", 
     cex = 1.5, col="blue")
#dev.off()
```


<p>&nbsp;</p>
### Plot Fig. 6.5: 1D EOF Errors by R code
```{r}
#Ms= 100 samples for North's rule of thumb
#install.packages("mvtnorm")
library(mvtnorm)
set.seed(112)
M=100 #M samples or M independent time steps
N=100 #N spatial locations
Mode=5 # 5 EOF modes to be considered

#Generate PCs
lam=c(10.0, 9.0, 4.1, 4.0, 2.0)
round(sqrt(2/M)*lam, digits = 2)
#[1] 1.41 1.27 0.58 0.57 0.28
univar =diag(rep(1,Mode)) #SD = 1
zeromean = rep(0, Mode) #mean = 0
rs <- rmvnorm(M, zeromean, univar) 
dim(rs) # 100 samples and 5 modes
#[1] 100    5
mean(rs[,1])
#[1] -0.02524037
var(rs[,1])
#[1] 0.9108917
t <- seq(0, 2*pi, len=M)
a51 <-rs[,1]*sqrt(1/M)
sum(a51^2)
#[1] 0.9026492 is the variance approximation

pcm=matrix(0, nrow=M, ncol=Mode)
for(m in 1:Mode){pcm[,m] = rs[,m]*sqrt(1/M)} 
dim(pcm) #random and independent PCs
#[1] 100   5
sum(pcm[,1]^2) #verify the normality of a PC
#[1] 1.021924

#Generate EOFs for spatial patterns
eof <- function(n, x) sqrt(2)* sin(n*x)*sqrt((1/N))
x <- seq(0, pi,len=N) #N locations within [0,1]
sum(eof(3,x)^2) #verify the normality of an EOF
#[1] 0.99
sum(eof(1,x)*eof(2,x)) #verify the EOF orthogonality
#[1] 3.035766e-18

eofm <- matrix(0, nrow=N, ncol=Mode)
for (m in 1:Mode){eofm[,m]=eof(m,x)}
dim(eofm) #eofm are the 5 modes of EOF data
#[1] 100   5 #100 spatial locations and 5 modes

#Generate the random data with given EOFs
Lam = diag(sqrt(lam)) #eigenvalue matrix
da <- eofm%*%Lam%*%t(pcm) #spectral decomposition
dim(da) #random data at 100 spatial locations
#[1]  100 100
svdda <- svd(da)
round((svdda$d[1:5])^2, digits =2)
#[1] 10.14  8.12  3.83  3.26  2.10

png(file="fig0605a.png", width=600,height=300)
k=1
sum(svdda$u[,k]^2)
#[1] 1
par(mar=c(4.5,4.7,2,0.2))
plot(x,-svdda$u[,k], type="l", ylim=c(-0.2,0.2),
     xlab="x", ylab=paste("EOF", k), 
     main="EOF Mode (Blue Curve) vs Exact Mode (Red Curve)",
     col="blue",
     cex.lab=1.4, cex.axis=1.4,cex.main=1.4)
lines(x, eof(k,x), col='red')
#dev.off()
```


<p>&nbsp;</p>
### Plot Fig. 6.6: R Code for the EOF4 Error
```{r}
k = 4 #Use the SVD result from 100 samples R = 100
plot(x,svdda$u[,k], type="l", ylim=c(-0.33,0.33),
     xlab="x", ylab="EOFs [dimensionless]", 
     main="EOF4 error (black) vs Exact EOF3 (Orange)",
     col="blue",
     cex.lab=1.4, cex.axis=1.4, cex.main=1.4)
legend(0.2,0.37, bty = "n", cex=1.4, text.col = 'blue',
       lwd=1.2,legend="Sample EOF4",col="blue")
lines(x, eof(k,x), col='red')
legend(0.2,0.41, bty = "n", cex=1.4, text.col = 'red',
       lwd=1.2,legend="True EOF4",col="red")
lines(x,svdda$u[,k] - eof(k,x), lwd=2, col='black'  )
legend(0.2,0.33, bty = "n", cex=1.4, text.col = 'black',
       lwd=2,legend="Sample EOF4 - True EOF4",col="black")
lines(x, -eof(3,x), col='orange')
legend(0.2,0.29, bty = "n", cex=1.4, text.col = 'orange',
       lwd=1.2,legend="True EOF3",col="orange")

```


<p>&nbsp;</p>
### Plot Fig. 6.7: R Code for 2D Sample EOFs
```{r}
library(mvtnorm)
dmvnorm(x=c(0,0))
M=100
N=100
Mode=5

#Generate PCs
lam=c(10.0, 9.0, 4.1, 4.0, 2.0)
univar =diag(rep(1,Mode))
zeromean = rep(0, Mode)
rs <- rmvnorm(M, zeromean, univar)
dim(rs)
#[1] 100    5
t <- seq(0, len=M)
a51 <-rs[,1]*sqrt(1/M)
pcm=matrix(0, nrow=M, ncol=Mode)
for(m in 1:Mode){pcm[,m] = rs[,m]*sqrt(1/M)} 
dim(pcm)
#[1] 100   5

#Generate true EOFs
eof <- function(n, x, y){(outer(sin(n*x),sin(n*y)))*(2/N)}
#eof <- function(n, x, y){outer(sin(n*x),sin(n*y))}
x = y <- seq(0,pi,len=100)
sum((eof(2,x, y))^2) #verify the normality
#[1] 0.9801

#Plot true EOF1 2D
png(file="fig0607a.png", width=200,height=200)
par(mar=c(2,2,0.5,0.5))
contour(x,y, eof(1,x,y))
text(1.5,1.5, 'EOF1', cex=2)
dev.off()

eofm <- matrix(0, nrow=N^2, ncol=Mode)
for (m in 1:Mode){eofm[,m]=c(eof(n=m,x,y))}
dim(eofm)
#[1] 10000   5 #10000 = 100*100
t(eofm)%*%eofm 
#approximately equal to I5 an identify matrix

#Generate the random data with given EOFs
Lam = diag(sqrt(lam))
da <- eofm%*%Lam%*%t(pcm)
dim(da)
#[1] 10000   100
svdda <- svd(da)

#Plot ssample EOF1-5 2D: R=100
#png(file="fig0607n.png", width=200,height=200)
par(mar=c(2,2,0.5,0.5))
k=5
contour(x,y, matrix(svdda$u[,k], ncol=100), 
        cex.lab=1.4, cex.main=1.4)
text(1.5,1.5, 'EOF5', cex=2)
#dev.off()
```
